Learning pleiotropic signatures to construct a causal map of the human phenome

### 2017-07-29 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George Davey Smith},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Learning pleiotropic signatures to construct a causal map of the human
phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George
Davey Smith}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 July 2017}

\usepackage{amsmath}

@@

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this. Here we show that the
mechanisms that underlie horizontal pleiotropy are numerous, and that
instrument selection will be increasingly liable to selecting invalid
instruments as GWAS sample sizes continue to grow. We have developed a
mixture of experts machine learning approach (MR-MoE 1.0) that improves
on both power and false discovery rates over all existing methods. Using
the approach, we systematically estimated the causal effects amongst
2407 phenotypes, generating a draft of the causal map of the human
phenome, but we offer caution that its interpretation is far from
straightforward.

@@

Mendelian randomization (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the exposure, and can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant represents
an independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) some violations of the assumptions of MR (most notably that the
SNP influences the outcome through no pathways other than the exposure)
can be evaluated through sensitivity analyses (3--7).

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes (8). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using only summary level data (known as two-sample MR, 2SMR)
(9). Here, in order to infer the causal effect of an exposure on an
outcome all that is required is an estimate of the genetic effects of
the instrumenting SNP on the exposure, and the corresponding estimate of
the effect on the outcome. This has two major advantages. First, GWAS
summary data is non-disclosive and often publicly available. Second,
causal inference can be made between phenotypes even if they have not
been measured in the same samples, limiting the breadth of possible
causal estimates only to the availability of GWAS summary data for the
traits in question (10).

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure. The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (3--6). It is considered best
practice to report estimates from all available methods as sensitivity
analyses when presenting causal estimates, however this strategy is not
necessarily optimal for several reasons. First, if different methods
disagree it is not possible to know which is correct because the true
nature of HP exhibited by the instruments is not known. Second, though
the IVW approach is most statistically powerful under no HP, it can have
high false negative or low true positive rates in the presence of HP
compared to other methods. Given that pleiotropy has been hypothesised
to be universal (11,12), defaulting to the IVW method in the first
instance and using other methods as sensitivity analyses may not be
appropriate. Third, the available methods do not cover all possible
models of HP, and therefore an automated method for instrument selection
may be necessary. Fourth, it could be of interest to make causal effect
estimates for thousands of traits, in which case a critical evaluation
of each causal estimate of interest may not be possible or convenient.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (10), we use these
new methods to construct a graph of millions of causal estimates. We
consider this to be a `draft' of the causal map of the human phenome,
but raise caution throughout that its interpretation is far from
straightforward

@@

The use of summary data in two-sample MR is described in detail
elsewhere (10). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of MR (that the instrument associates with the exposure) is
satisfied. We collect their effect sizes, standard errors and effect
alleles for the association with the exposure trait. These can be
obtanied manually from complete summary data or from curated lists of
significant GWAS associations. Next, the effects of those SNPs on the
outcome need to be obtained, typically necessitating complete summary
data because it is unlikely that these SNPs will have reached
genome-wide significance (and therefore be present in curated
catalogues). At its most simple implementation, the regression of the
SNP-exposure effect sizes against the SNP-outcome effect sizes, with
greater weight afforded to those SNPs with smaller SNP-outcome standard
errors, provides the estimate of the causal effect of the exposure on
the outcome. This is known as the inverse-variance weighted (IVW)
method.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there is sufficient summary data available.
Supplementary table 1 provides a list of all traits that have available
GWAS summary data, indicating if they have complete summary data (in
which case they can be used as both exposure traits and outcome traits),
or if they only have significant associations (in which they can only be
used as exposure traits).

@@

\textbf{Mean-based methods:} Here we consider four nested models (5).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression and
not introducing bias. Fixed effects Egger regression (3) relaxes the HP
assumption further by allowing a non-zero intercept which essentially
allows horizontal pleiotropy to be directional, where it systematically
influences the outcome in a specific direction. Random effects Egger
regression allows heterogeneity around the directional HP, as long as
the HP effects are not correlated with the SNP-exposure effects (also
known as the INSIDE assumption).

The Rucker framework (13), adapted to MR (5) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of instruments) can be used to obtain a sampling distribution for the
model estimate amongst these four variations. Using 1000 rounds of
jackknife estimates, we can obtain a final estimate using the mean or
the median of the distribution. We only use the jackknife approach for
associations where there are 15 or more variants to avoid combinatorial
saturation. The four nested models plus the three Rucker estimates
provide seven mean based estimators.

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (4). This has the advantage
that up to half the instruments can be invalid, and the estimate will
remain unbiased. Developing the approach further to allow stronger
instruments to contribute more towards the estimate can be obtained by
obtaining the median of the weights of each instrument. The penalised
weighted median estimator introduces a further weight to the
instruments, penalising any instrument that contributes substantially
towards the heterogeneity statistic. Together, this provides three
median based estimators.

\textbf{Mode-based methods:} Similar to the median, the mode based
estimator clusters the instruments into groups based on similarity of
causal effects, and returns the final causal effect estimate based on
the cluster that has the largest number of instruments (6). This can be
extended in two ways, first by weighting the strength of each cluster by
the inverse of the variance of the SNP-outcome associations in the
cluster; second by weighting the strength of each cluster by the inverse
of the variance of the SNP-exposure associations. When weighting by the
SNP-outcome standard errors we term this a weighted analysis. When
\emph{not} weighting by the exposure standard errors we term this the
`no measuremenent error in the exposure (NOME)' assumption. Together
this provides four mode-based estimators.

@@

The simplest approach to selecting instruments for performing MR is to
take take SNPs that have been declared significant in the published GWAS
for the exposure. This typically involves obtaining SNPs that surpass
\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
then replicating in an independent sample. These results are often
recorded in public GWAS catalogs. Alternatively the clumping procedure
can be performed using complete summary data in MR-Base. We call this
the ``top hits'' strategy.

@@

An approach to inferring the causal direction between phenotypes was
developed recently (7), using the following basic premise. If trait
\(A\) causes trait \(B\) then

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously. However, we use it here as an inexpensive and
approximate method to identify variants that are likely to be invalid.
Steiger's Z-test of correlated correlations (14) can be used to formally
test the extent to which the two correlations are statistically
different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). In this case the Steiger test applied to each
variant in turn and we exclude any \(g_{A}\) for which the Steiger
Z-test has \(p > 0.05\), indicating that it is unlikely to primarily
associate with \(B\) relative to \(A\). Similarly, for SNPs that
influence confounders of \(A\) and \(B\) or those variants that exhibit
horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant.

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
\(p\) is the allele frequency of the SNP in the population (15). \(p\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\). We
exclude any SNPs that do not have the correct causal direction for the
hypothesis, and do not have Steiger test \(p < 0.05\).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). In the context of this
analysis, each method is considered to be an `expert', taking a set of
SNP-exposure and SNP-outcome effect sizes and their standard errors as
inputs. Our objective is to select the expert most likely to be correct
for a specific MR analysis.

The mixture of experts method is a machine learning approach (16) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that dataset.
The metrics are a mixture of regression diagnostics and are described in
Supplementary table 3.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated datasets. We trained the gating function using random forest
learning algorithms that seek to identify sectors of the parameter space
that are most likely to return accurate estimates for a particular
expert. Figure 2a illustrates how the gating function is trained using
simulated data. New datasets can then be applied to the trained mixture
of experts (MoE) model (Figure 2b). There are countless ways to
implement this approach, we call this implementation MR-MoE 1.0.

@@

The MoE is trained using datasets generated from simulations (Figure
2a). A \emph{dataset} is the minimum data required to perform 2SMR
analysis - four columns comprising the SNP-exposure and SNP-outcome
effects and standard errors, and rows corresponding to each SNP that is
used as an instrument for the exposure. This \emph{dataset} can be fed
into any of the 28 experts to obtain MR causal effects. The simulations
used to generate these datasets seek to cover a range of pleiotropic
scenarios, including where some proportion of SNPs exhibit directional
or balanced horizontal pleiotropy, or where SNPs influencing confounding
variables.

@@

Some proportion \(s_x \sim U(0,1)\) of \(g_x\) SNPs and some proportion
\(s_y \sim U(0,1)\) of \(g_y\) SNPs exhibit horizontal pleiotropy with
effects sampled using

@@

\[
    O_{w,d} = 
\begin{cases}
    1,   & \text{if } \beta_{x,y} \neq 0 \text{and } p_{m,d} < 0.01\\
    1,   & \text{if } \beta_{x,y} = 0 \text{and } p_{m,d} > 0.1 \\
    0,   & \text{otherwise}
\end{cases}
\]

@@

For each training dataset we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated dataset. For each of our 28 methods, we need to create a model
that predicts the performance of the method based on metrics generated
from a dataset. To do this, for each method we train random forest
decision trees to predict that method's performance using the dataset's
metrics. The random forest approach is well suited to this problem
because there are likely to be non-continuous combinations of different
metrics that improve on prediction over, for example, a simple linear
model that does not learn about interactions.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{amsmath}

@@

\setlength{\droptitle}{-2em}
  \title{Learning pleiotropic signatures to construct a causal map of the human
phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{29 July 2017}

@@

Gibran Hemani\(^1\), Jack Bowden\(^1\), Philip Haycock\(^1\), Jie
Zheng\(^1\), Oliver Davis\(^1\), Peter Flach, Tom Gaunt\(^{1,*}\),
George Davey Smith\(^{1,*}\)

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
  School of Social and Community Medicine, Bristol, UK
\end{enumerate}

Correspondence to:
\href{mailto:g.hemani@bristol.ac.uk}{\nolinkurl{g.hemani@bristol.ac.uk}}

* Equal contribution

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this. Here we show that the
mechanisms that underlie horizontal pleiotropy are numerous, and that
instrument selection will be increasingly liable to selecting invalid
instruments as GWAS sample sizes continue to grow. We have developed a
mixture of experts machine learning approach (MR-MoE 1.0) that improves
on both power and false discovery rates over unselective use of any
existing methods. Using the approach, we systematically estimated the
causal effects amongst 2407 phenotypes, generating a working draft of
the causal map of the human phenome. Almost 90\% of causal estimates
indicated some level of horizontal pleiotropy. All results are organised
into a publicly available graph database (\url{http://eve.mrbase.org}).

@@

Mendelian randomization (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the if it only influences the outcome through the exposure
(the exclusion restriction). It can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant represents
an independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) potential bias arising from the exclusion restriction principal
can be detected or corrected by evaluating the consistency of effects
across instruments (3--8).

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes (9). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using only summary level data (known as two-sample MR, 2SMR)
(10). Here, in order to infer the causal effect of an exposure on an
outcome all that is required is an estimate of the genetic effects of
the instrumenting SNP on the exposure, and the corresponding estimate of
the effect on the outcome. This has three major advantages. First, GWAS
summary data are non-disclosive and often publicly available. Second,
causal inference can be made between phenotypes even if they have not
been measured in the same samples, limiting possible causal estimates
only by the availability of GWAS summary data for the traits in question
(11). Third, we can view the causal inference problem through the simple
and widely understood prism of a meta-analysis.

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure (2). The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (3--6,8,12). It is considered best
practice to report estimates from all available methods as sensitivity
analyses when presenting causal estimates because different pleiotropic
models can then be evaluated. However it is also desirable to
consolidate across methods under some circumstances. First, it could be
of interest to make causal effect estimates for thousands of traits, in
which case a critical evaluation of each causal estimate of interest may
not be possible or convenient. Second, though the IVW approach is most
statistically powerful under no HP, it can have high false negative or
low true positive rates in the presence of HP compared to other methods.
Pleiotropy has been hypothesised to be universal (13), though the degree
to which this may be the case is contested (14,15), hence defaulting to
the IVW method in the first instance and using other methods as
sensitivity analyses may not be appropriate. Third, if different methods
disagree it is not possible to know which is correct because the true
nature of HP exhibited by the instruments is not known. Fourth, the
available methods do not cover all possible models of HP, and therefore
an automated method for instrument selection may be necessary.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (11), we use these
new methods to construct a graph of millions of causal estimates.
Motivated by the recent avalanche (16) of 2SMR publications, often with
repeating (17,18) or contradicting (19,20) conclusions despite using the
same data, we developed a graph database to represent these estimates in
a consistent manner. We consider this to be a `working draft' of the
causal map of the human phenome, but raise caution throughout that its
interpretation is far from straightforward, and that future corrections
and refinements will inevitably follow as data grows and 2SMR methods
evolve.

@@

The use of summary data in two-sample MR is described in detail
elsewhere (11). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of MR (that the instrument associates with the exposure) is
generally satisfied. We collect their effect sizes, standard errors and
effect alleles for the association with the exposure trait. These can be
obtained manually from complete summary data or from curated lists of
significant GWAS associations. Next, the effects of those SNPs on the
outcome need to be obtained, typically necessitating complete summary
data because these SNPs will generally have not have reached genome-wide
significance for other traits (and therefore be present in curated
catalogues). At its most simple implementation, the regression of the
SNP-exposure effect sizes against the SNP-outcome effect sizes, with
greater weight afforded to those SNPs with smaller SNP-outcome standard
errors, provides the estimate of the causal effect of the exposure on
the outcome. This is known as the inverse-variance weighted (IVW)
method.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there are sufficient summary data available.
Supplementary table 1 provides a list of all traits that have available
GWAS summary data, indicating if they have complete summary data (in
which case they can be used as both exposure traits and outcome traits),
or if they only have significant associations (in which they can only be
used as exposure traits).

@@

\textbf{Mean-based methods:} Here we consider four nested models (5).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression and
not introducing bias. Fixed effects Egger regression (3) relaxes the HP
assumption further by allowing a non-zero intercept which essentially
allows horizontal pleiotropy to be directional, where it systematically
influences the outcome in a specific direction. Random effects Egger
regression allows heterogeneity around the directional HP (5), as long
as the HP effects are not correlated with the SNP-exposure effects (also
known as the INSIDE assumption)(3).

The Rucker framework (21), adapted to MR (5) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of the complete set of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations. Using
1000 rounds of jackknife estimates, we can obtain a final estimate using
the mean or the median of the distribution. We only use the jackknife
approach for associations where there are 15 or more variants in order
to avoid saturating the possible number of instrument combinations.

The four nested models (IVW fixed effects, IVW random effects, Egger
fixed effects, Egger random effects) plus the three Rucker estimates
(point estimate, mean of the jackknife, median of the jackknife) provide
seven mean based estimators.

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (4,22). This has the
advantage that up to half the instruments can be invalid, and the
estimate will remain unbiased. Developing the approach further to allow
stronger instruments to contribute more towards the estimate can be
obtained by obtaining the median of the weights of each instrument. The
penalised weighted median estimator introduces a further weight to the
instruments, penalising any instrument that contributes substantially
towards the heterogeneity statistic. Together, this provides three
median based estimators. Other estimation strategies not considered
here, such as LASSO regression, have also been developed for the
situation where at least half of the instruments are valid (23).

\textbf{Mode-based methods:} Similar to the median, the mode based
estimator clusters the instruments into groups based on similarity of
causal effects, and returns the final causal effect estimate based on
the cluster that has the largest number of instruments (6). There are
four implementations of this method: the simple and the weighted mode,
each weighted with or without the assumption of no measurement error in
the exposure estimates (NOME). The simple mode is the unweighted mode of
the empirical density function of causal estimates, whereas the weighted
mode is weighted by the inverse variance of the outcome effect. The
bandwidth parameter was set to 1 by default.

@@

The simplest approach to selecting instruments for performing MR is to
take SNPs that have been declared significant in the published GWAS for
the exposure. This typically involves obtaining SNPs that surpass
\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
then replicating in an independent sample. These results are often
recorded in public GWAS catalogs. Alternatively the clumping procedure
can be performed using complete summary data in MR-Base (11). We call
this the ``top hits'' strategy.

@@

An approach to inferring the causal direction between phenotypes
recently developed (7) uses the following basic premise. If trait \(A\)
causes trait \(B\) then

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously. However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid. Steiger's Z-test of correlated correlations (24) can be
used to formally test the extent to which the two correlations are
statistically different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(8,12,25), based on the notion that outlying instruments are likely to
be exhibiting horizontal pleiotropy. Conversely, we primarily developed
Steiger filtering to identify those instruments that are likely to be
arising due to reverse cause.

The Steiger test is applied to each variant in turn and we exclude any
\(g_{A}\) for which \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\), indicating
that it is unlikely to primarily associate with \(B\) relative to \(A\).
Similarly, for SNPs that influence confounders of \(A\) and \(B\) or
exhibit horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant. Hence we also exclude any \(g_{A}\) for which the Steiger
Z-test has \(p > 0.05\).

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
\(p\) is the allele frequency of the SNP in the population (26). \(p\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\). We
exclude any SNPs that do not have the correct causal direction for the
hypothesis, and do not have Steiger test \(p < 0.05\).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). In the context of this
analysis, each method is considered to be an `expert', taking a set of
SNP-exposure and SNP-outcome effect sizes and their standard errors as
inputs. Our objective is to select the expert most likely to be correct
for a specific MR analysis. A overview of the approach is shown in
Figure 2.

The mixture of experts method is a machine learning approach (27) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that dataset.
The metrics are a mixture of regression diagnostics and are described in
Supplementary table 3.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated datasets. We trained the gating function using random forest
learning algorithms that seek to identify sectors of the parameter space
that are most likely to return accurate estimates for a particular
expert. Figure 2a illustrates how the gating function is trained using
simulated data. New datasets can then be applied to the trained mixture
of experts (MoE) model (Figure 2b). We call this implementation MR-MoE
1.0.

It should be noted that in the original hierarchical mixture of experts
approach of (27) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (28--30) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the datasets.

@@

The MoE is trained using datasets generated from simulations (Figure
2a). A \emph{dataset} is the minimum data required to perform 2SMR
analysis - four columns comprising the SNP-exposure and SNP-outcome
effects and standard errors, and rows corresponding to each SNP that is
used as an instrument for the exposure. This \emph{dataset} can be fed
into any of the 28 experts to obtain MR causal effects. The simulations
used to generate these datasets seek to cover a range of pleiotropic
scenarios, including where some proportion of SNPs exhibit directional
or balanced horizontal pleiotropy, or where SNPs influence confounding
variables.

@@

Some proportion \(floor(MS_{x}/M)\) of \(g_x\) SNPs and
\(floor(MS_{y}/M)\) of \(g_y\) SNPs, where \(s_x\) and
\(s_y \sim U(0,1)\), exhibit horizontal pleiotropy with effects sampled
using

@@

\[
    O_{w,d} = 
\begin{cases}
    1,   & \text{if } \beta_{x,y} \neq 0 \text{ and } p_{m,d} < 0.01\\
    1,   & \text{if } \beta_{x,y} = 0 \text{ and } p_{m,d} > 0.1 \\
    0,   & \text{otherwise}
\end{cases}
\]

where \(p_{m,d}\) is the p-value for method \(m\) on dataset \(d\).

@@

For each training dataset we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated dataset. For each of our 28 methods, we need to create a model
that predicts the performance of the method based on metrics generated
from a dataset. To do this, for each method we train a random decision
forest to predict that method's performance using the dataset's metrics.
The random forest approach is well suited to this problem because there
are likely to be non-continuous combinations of different metrics that
improve on prediction over, for example, a simple linear model that does
not learn about interactions.

As a simple hypothetical example - if a dataset has a single outlier but
is otherwise exhibiting no heterogeneity then the following methods
could arguably perform well:
